Comparing vector-based and ACT-R memory models using large-scale datasets:\protect\\User-customized hashtag and tag prediction on Twitter and StackOverflow

### 2014-10-23 writeups/dissertation/dissertation.tex
foo

I would also like to mention a few open-sourced software packages that were heavily leveraged to do this research:
The majority of the modeling and analysis was done with the R statistical computing environment \parencite{RCoreTeam2014} and two important R packages:
data.table \parencite{Dowle2014} was leveraged to implement the models efficiently, and ggplot2 \parencite{Wickham2009} was used to produce the visualizations.

@@

Also, studying user-based tag generation on these sites has relevant real-world application.
Models that can accurately predict the tags that users will generate on Twitter and StackOverflow can be used as the foundation for recommendations systems on these popular sites.
These systems can also help newer users by recommending proper tags for their newest content.
On the StackOverflow site, for example, experts often subscribe to specific hashtags that interest them, and receive a daily digest of posts that are tagged with those hashtags.
Helping the user properly tag posts on StackOverflow ensures that right community sees the post, which greatly increases the chance that the question on the post will be answered quickly and correctly.
For Twitter, hashtags represent streams of information that are possibly interesting to the user.
Having a recommendation system that can suggest relevant hashtags to the user provides a way for the user to connect to and discover new information streams that interest them.
Also, by suggesting relevant hashtags the system can help make sure a user's content is seen by other users who are interested in the same type of information.

@@

The fan effect describes a behavior of human memory where memory items are more difficult to retrieve if they have occurred in a larger number of different contexts.
\textcite{Rutledge2008} used \textcite{Anderson1974}'s classic experimental paradigm to test the fan effect.
In the task, participants are exposed to a set of person and place pairs (e.g., ``the hippie is in the park'').
They are later asked to recall if a particular person and place pair was presented to them before.
The main manipulation is the ``fan'' of the person and place for each pair.
The fan is defined as the number of different contexts that each item in the pair is presented in.
For example, if items were presented to the participant where the ``hippie'' was both in the ``park'' and in the ``store,'' then ``hippie'' would have a fan of two.
Recall accuracy decrease and time to recall increase as the sum total fan of the person and pair increases.

@@

To better illustrate the dynamics of each user's tagging behavior,
profile plots for a single user for both StackOverflow and Twitter are included in Figure \ref{figPriorHTByTime}.
The specific value of the hashtag used for both plots is a nominal variable and not important for this analysis.
What is interesting is the large tag space for these users.
There are bands of commonly-used tags for each user, but there are also many instances where new tags are used, and a large percentage of tags used are not in the most common bands.
For example, the StackOverflow user in Figure \ref{figPriorHTByTime} has a pronounced band where the same hashtag is chosen repeatedly over time. 
However there are many instances of other tags being used in parallel to the tag that is producing the band (that is repeatedly being used). 
The Twitter user has a few clusters where the same hashtag is chosen repeatedly but only for a brief period of time.
As time progresses, a band of choosing another hashtag is observed, and then the user switches to a new hashtag after that.
More recent posts for the Twitter user are more varied, and there is no longer a single hashtag that the user chooses repeatedly.
This helps illustrate how challenging modeling this task can be.
The model's goal is to predict at each instance what tag will be chosen, but to do so the model must predict the correct few tags from a much larger set of tags whose frequency of occurrence is changing over time.

@@

@@

\begin{figure}[!htbp]
  \scalebox{.8}{\resizebox{\linewidth}{!}{\includegraphics{HTMByTime-520957-crop.pdf}}}
  \caption{Model tagging profile of a single user on StackOverflow}
  \label{figPriorHTMByTimeSO}
\end{figure}

\begin{figure}[!htbp]
  \scalebox{.8}{\resizebox{\linewidth}{!}{\includegraphics{HTMByTime-fashionista_com-crop.pdf}}}
  \caption{Model tagging profile of a single user on Twitter}
  \label{figPriorHTMByTimeT}
\end{figure}

@@

Finally, model accuracy was analyzed across all dataset slices for StackOverflow and Twitter.
This was done by examining model accuracy at a specific decay rate value for each model.
Accuracy was assessed at each model's previously-determined best-fit values (\num{.43} and \num{.80} for the optimized learning and standard form), and the standard form's default value (\num{0.5}).
Since decay rate values were explored at \num{0.1} increments between the 0 and 1 range, the decay rate values used to examine model accuracy were rounded to the nearest decay rate value that was included in the run.
So a best-fit value of \num{.4} was used for the optimized learning model instead of the \num{.43} average found from the dataset slices.
Accuracy for both sites is included in Figure \ref{figPriorAcc}. 
Aggregate accuracy was computed by taking the mean of every user's model accuracy across all users in all dataset slices.

@@

ACT-R sets the default decay rate value to \num{0.5}, and the best-fit value for the standard form is higher (\num{0.8}) for these datasets.
The figure shows that although model accuracy does improve for the standard form when increasing the decay rate parameter from \num{0.5} to the optimal value,
that percentage improvement is not as large as the improvement gained from using the standard form compared to the optimized learning form.
Nonetheless, model accuracy is highest when both the standard form is used and the decay rate value is increased.

@@

In order to best combine these predictors, a logistic regression statistical technique similar to \textcite{Stanley2013} was used to determine the most optimal weights for each term.
\textcite{Stanley2013} weighted each term in the total activation equation Table \ref{tabModACTRModel}) to maximize model accuracy.
\citeauthor{Stanley2013} found that the optimal weights for the StackOverflow model for the prior term, title context term, and body context term were \num{.62}, \num{.93}, and \num{1.75} respectively.

The optimal weights for each model component are computed in the same manner as \textcite{Stanley2013}:
When a user creates a tag for a post, a retrieval request for the model is made.
For each request, the model returns a set of tags (all tags that have been observed for that user in the past) and activations associated with each tag.
An activation value is computed for each model component (e.g., prior tag use, context) for each tag.
These values are recorded, and tag instances that match what the author actually tagged are marked as a 1, while all others are marked with a 0.
This process is repeated across all posts where a user chooses a tag, and across all users in the dataset.
Once all of the results are aggregated, a logistic regression is run where the optimal weights for each model component are found that maximize the model's ability to correctly label tags as 0 or 1 based on total activation.

Model accuracy is evaluated by rank ordering all recorded tags by model activation using the optimal weights,
asking the model to tag the top N posts with a 1 (where N is the total number of recorded tagging instances in the dataset sample),
and then comparing the model's chosen tags with the used tags for each post (i.e., looking at the ``hits'').
This is equivalent to throttling the threshold in the logistic regression such that the total number of observations labeled as a 1 match the total number of recorded instances of tagging in the dataset.

@@

One simple and likely explanation is that there are posts where a large set of tags have high activation,
but the author is choosing to use only a fixed small set of the top tags for each post, regardless of the activation values of the top tags.
The author determines the number of tags to use for a post based on competing goals of tagging posts an not overly saturating a post with tags.
For example, the author may be writing a question on StackOverflow about \emph{javascript}, and a large set of related tags are retrieved from declarative memory that are all above the retrieval threshold
(e.g., \emph{javascript}, \emph{dom}, \emph{svg}, \emph{html}, \emph{jquery}, \emph{css}),
but the author only chooses to use the tags \emph{javascript}, \emph{dom}, and \emph{svg}, since he has decided to tag this post with three tags, and those are the three tags with the highest activation.
Adding the offset term into the model is a way to implement a simple version of this strategy where the author is actively choosing to use only a small set of tags for each post.
Also, this type of strategic behavior of limiting the total number of chosen tags could be easily implemented in the production system of ACT-R.

@@

Another interesting result was how the Bayesian model actually performed worse when stop words were removed based on frequency compared to no stop word removal.
This was not the case for the random permutation model in this study, was not the case when \textcite{Sahlgren2008} used frequency filtering on the random permutation model,
and is most likely never the case for the random permutation model.
Model accuracy most likely decreases for the Bayesian model because this model already has a normalization component built in that the random permutation model does not have.
When computing activation for the random permutation model, the correlation is calculated directly on a matrix of counts of co-occurrences for words and tags.
For the Bayesian model this count matrix is converted to a log-odds ($S_{ji}$) matrix,
where both the number of observations for the word and the tag are normalized when computing the value for each cell (see the $S_{ji}$ computation in Table \ref{tabModACTRModel}).
So it is likely the case that a form of frequency attenuation is already built into the Bayesian model.
Therefore, adding a frequency filter on top of this does nothing to improve accuracy and only removes predictive information.

@@

The transition point where accuracy for the Bayesian model outperformed the random permutation model was much higher for StackOverflow than Twitter.
This is most likely because StackOverflow has a almost \num{100} times more unique tags than the Twitter popular-hashtags dataset.
The number of words in a StackOverflow post is also much higher than Twitter, which should have lowered the transition point.
However, the effect of words in a post must be less than the effect of number of unique tags, as the transition point was still much higher for StackOverflow given that it should have lowered due to post size.

@@

Another parameter in the random permutation model is the number of rows used to represent the all of the different words.
Each word's environment vector is represented by two randomly placed ones and two negative ones across the rows.
All that is required is that the four positions and signs of each of these environment vectors is unique.
But this means multiple environment vectors can overlap when looking at a single row.
This is a lossy form of compression, and allows the number of rows to represent the words to be much less than the total number of words.
As the number of rows decreases, the amount of cross talk between the environment vectors increases, compression increases, and the amount of information stored decreases.
As a third architectural concern, the effect of compression for the random permutation model was examined for Twitter and StackOverflow.

@@

\begin{figure}[!htbp]
  \scalebox{.8}{\resizebox{\linewidth}{!}{\includegraphics{compareMeanDV-acc-dimBySizeSO-crop.pdf}}}
  \caption{Effect of compression on random permutation model for StackOverflow}
  \label{figContextDimSO}
\end{figure}

@@

\begin{figure}[!htbp]
  \scalebox{.8}{\resizebox{\linewidth}{!}{\includegraphics{compareMeanDV-acc-dimBySizeT-crop.pdf}}}
  \caption{Effect of compression on random permutation model for Twitter}
  \label{figContextDimT}
\end{figure}

@@

The original BEAGLE model \parencite{Jones2007} actually showed only minimal improvement when incorporating word order as well.
In absolute terms, this model only improved by 2 percentage points in accuracy when trained on the TASA and tested on the TOEFL.
\textcite{Sahlgren2008}'s results for word order for the random permutation model are similar.
At a row size of around \num{2000}, model accuracy improves by less than a percentage point when order information is incorporated.
The results for larger row sizes are somewhat inconsistent, since the full model (order and orderless) sometimes underperforms the order-only model,
which should never happen since the order-only model is a subset of the full model.
\textcite{Sahlgren2008} most likely did not use a regression technique to optimize the weights for each of the order and orderless model terms.
So although incorporating word order into the random permutation model does seem to slightly improve performance,
the strength of the model is more likely to be in how efficiently it can represent the co-occurrence matrix rather than its ability to include word order in the representation.

@@

This raises an important applied question: Exactly how much past user behavior is enough for a site like StackOverflow to benefit from adding past user behavior as an additional predictor.
When the model predicts tags from a random sample of StackOverflow users, adding the prior component to the model does not significantly increase model accuracy compared to using context alone.
However, when the model predicts tags from a set of users that ask the most questions on the site, adding the prior does significantly increase accuracy.
For the StackOverflow dataset, a user asks on average 5 questions and adds 3 tags to a post.
So the model uses approximately 15 prior observations when computing the prior component for a randomly-sampled user on the site.
For the popular-users dataset used to generate the results in Figure \ref{figPUAddPrior}, a user asked 180 questions on average.
So for this dataset, the model uses approximately 540 prior observations when computing the prior activation.
So for the StackOverflow site, the prior component begins to generate unique predictive variance above and beyond the context component somewhere between when a user has generated 15 and 540 prior observations.
This result will be different for other sites that have collected a different amount of contextual information.
For sites like StackOverflow where the contextual component is already a strong predictor, it will take a larger number of prior observations before adding the prior component becomes beneficial.
Alternatively for sites like Twitter where contextual information is much more noisy and not as strong of a predictor, it will take fewer observations before adding the prior becomes beneficial.

@@

Given that the storage space required to represent the co-occurrences for the random permutation is dramatically smaller than the uncompressed Bayesian model,
and that the compression is lossy and essentially random (i.e., no SVD),
it is quite interesting that the random permutation model performs nearly as well as the full Bayesian model for this task.

>>>>

I would also like to thank the creators of a few open-sourced software packages that were heavily leveraged to do this research:
The majority of the modeling and analysis was done with the R statistical computing environment \parencite{RCoreTeam2014} and two important R packages:
data.table \parencite{Dowle2014} was leveraged to implement the models efficiently, and ggplot2 \parencite{Wickham2009} was used to produce the visualizations.

@@

Also, studying user-based tag generation on these sites has relevant real-world application.
Models that can accurately predict the tags that users will generate on Twitter and StackOverflow can be used as the foundation for recommendations systems on these popular sites.
These systems can also help newer users by recommending proper tags for their newest content.
On the StackOverflow site, for example, experts often subscribe to specific tags that interest them, and receive a daily digest of posts that tagged with those tags.
Helping the user properly tag posts on StackOverflow ensures that right community sees the post, which greatly increases the chance that the question on the site will be answered quickly and correctly.
For Twitter, hashtags represent streams of information that are possibly interesting to the user.
Having a recommendation system that can suggest relevant hashtags to the user provides a way for the user to connect to and discover new information streams that interest them.
Also, by suggesting relevant hashtags the system can help make sure a user's content is seen by other users who are interested in the same type of information.

@@

The fan effect describes a behavior of human memory where memory items are more difficult to retrieve if they have occurred in a larger number of different contexts.
\textcite{Rutledge2008} used \textcite{Anderson1974}'s classic experimental paradigm to test the fan effect.
In the task, participants are exposed to a set of person and place pairs (e.g., ``the hippie is in the park'').
They are later asked to recall if a particular person and place pair was presented to them before.
The main manipulation is the ``fan'' of the person and place for each pair.
The fan is defined as the number of different contexts that each item in the pair is presented in.
For example, if items were presented to the participant where the ``hippie'' was both in the ``park'' and in the ``store,'' then ``hippie'' would have a fan of two.
Recall accuracy decrease and time to recall increase as the sum total fan of the person and place increases.

@@

To better illustrate the dynamics of each user's tagging behavior,
profile plots for a single user for both StackOverflow and Twitter are included in Figure \ref{figPriorHTByTime}.
The specific value of the hashtag used for both plots is a nominal variable and not important for this analysis.
What is interesting is the large tag space for these users.
There are bands of commonly-used tags for each user, but there are also many instances where new tags are used, and a large percentage of tags used are not in the most common bands.

@@

For example, the StackOverflow user in Figure \ref{figPriorHTByTime} has a pronounced band where the same hashtag is chosen repeatedly over time. 
However there are many instances of other tags being used in parallel to the tag that is producing the band (that is repeatedly being used). 
The Twitter user has a few clusters where the same hashtag is chosen repeatedly but only for a brief period of time.
As time progresses, a band of choosing another hashtag is observed, and then the user switches to a new hashtag after that.
More recent posts for the Twitter user are more varied, and there is no longer a single hashtag that the user chooses repeatedly.
This helps illustrate how challenging modeling this task can be.
The model's goal is to predict at each instance what tag will be chosen, but to do so the model must predict the correct few tags from a much larger set of tags whose frequency of occurrence is changing over time.

@@

\begin{figure}[!htbp]
  \scalebox{.7}{\resizebox{\linewidth}{!}{\includegraphics{HTMByTime-520957-crop.pdf}}}
  \caption{Model tagging profile of a single user on StackOverflow}
  \label{figPriorHTMByTimeSO}
\end{figure}

\begin{figure}[!htbp]
  \scalebox{.7}{\resizebox{\linewidth}{!}{\includegraphics{HTMByTime-fashionista_com-crop.pdf}}}
  \caption{Model tagging profile of a single user on Twitter}
  \label{figPriorHTMByTimeT}
\end{figure}

@@

Finally, model accuracy was analyzed across all dataset slices for StackOverflow and Twitter.
This was done by examining model accuracy at a specific decay rate value for each model.
Accuracy was assessed at each model's previously-determined best-fit values (\num{.43} and \num{.80} for the optimized learning and standard form) and the standard form's default value (\num{0.5}).
Since decay rate values were explored at \num{0.1} increments between the 0 and 1 range, the decay rate values used to examine model accuracy were rounded to the nearest decay rate value that was included in the run.
So a best-fit value of \num{.4} was used for the optimized learning model instead of the \num{.43} average found from the dataset slices.
Accuracy for both sites is included in Figure \ref{figPriorAcc}. 
Aggregate accuracy was computed by taking the mean of every user's model accuracy across all users in all dataset slices.

@@

ACT-R sets the default decay rate value to \num{0.5}, and the best-fit value for the standard form is higher (\num{0.8}) for these datasets.
The figure shows that although model accuracy does slightly improve for the standard form when increasing the decay rate parameter from \num{0.5} to the optimal value,
that percentage improvement is not as large as the improvement gained from using the standard form compared to the optimized learning form.
Nonetheless, model accuracy is highest when both the standard form is used and the decay rate value is increased.

@@

In order to best combine these predictors, a logistic regression statistical technique similar to \textcite{Stanley2013} was used to determine the most optimal weights for each term.
\textcite{Stanley2013} weighted each term in the total activation equation (Table \ref{tabModACTRModel}) to maximize model accuracy.
\citeauthor{Stanley2013} found that the optimal weights for the StackOverflow model for the prior term, title context term, and body context term were \num{.62}, \num{.93}, and \num{1.75} respectively.

The optimal weights for each model component were computed in the same manner as \textcite{Stanley2013}:
When a user creates a tag for a post, a retrieval request for the model is made.
For each request, the model returns a set of tags (all tags that have been observed for that user in the past) and activations associated with each tag.
An activation value is computed for each model component (e.g., prior tag use, context) for each tag.
These values are recorded, and tag instances that match what the author actually tagged are marked as a 1, while all others are marked with a 0.
This process is repeated across all posts where a user chooses a tag, and across all users in the dataset.
Once all of the results are aggregated, a logistic regression is run where the optimal weights for each model component are found that maximize the model's ability to correctly label tags as 0 or 1 based on total activation.

Model accuracy was evaluated by rank ordering all recorded tags by model activation using the optimal weights,
asking the model to tag the top N posts with a 1 (where N is the total number of recorded tagging instances in the dataset sample),
and then comparing the model's chosen tags with the used tags for each post (i.e., looking at the ``hits'').
This is equivalent to throttling the threshold in the logistic regression such that the total number of observations labeled as a 1 match the total number of recorded instances of tagging in the dataset.

@@

One simple and likely explanation is that there are posts where a large set of tags have high activation,
but the author is choosing to use only a fixed small set of the top tags for each post, regardless of the activation values of the top tags.
The author determines the number of tags to use for a post based on competing goals of tagging posts and not overly saturating a post with tags.
For example, the author may be writing a question on StackOverflow about \emph{javascript}, and a large set of related tags are retrieved from declarative memory that are all above the retrieval threshold
(e.g., \emph{javascript}, \emph{dom}, \emph{svg}, \emph{html}, \emph{jquery}, \emph{css}),
but the author only chooses to use the tags \emph{javascript}, \emph{dom}, and \emph{svg}, since he has decided to tag this post with three tags, and those are the three tags with the highest activation.
Adding the offset term into the model is a way to implement a simple version of this strategy where the author is actively choosing to use only a small set of tags for each post.
Also, this type of strategic behavior of limiting the total number of chosen tags could be easily implemented in the production system of ACT-R.

@@

Another interesting result is how the Bayesian model actually performed worse when stop words were removed based on frequency compared to no stop word removal.
This was not the case for the random permutation model in this study, was not the case when \textcite{Sahlgren2008} used frequency filtering on the random permutation model,
and is most likely never the case for the random permutation model.
Model accuracy most likely decreases for the Bayesian model because this model already has a normalization component built in that the random permutation model does not have.
When computing activation for the random permutation model, the correlation is calculated directly on a matrix of counts of co-occurrences for words and tags.
For the Bayesian model this count matrix is converted to a log-odds ($S_{ji}$) matrix,
where both the number of observations for the word and the tag are normalized when computing the value for each cell (see the $S_{ji}$ computation in Table \ref{tabModACTRModel}).
So it is likely the case that a form of frequency attenuation is already built into the Bayesian model.
Therefore, adding a frequency filter on top of this does nothing to improve accuracy and only removes predictive information.

@@

The transition point where accuracy for the Bayesian model outperformed the random permutation model was much higher for StackOverflow than Twitter.
This is most likely because StackOverflow has almost \num{100} times more unique tags than the Twitter popular-hashtags dataset.
The number of words in a StackOverflow post is also much higher than Twitter, which should have lowered the transition point.
However, the effect of words in a post must be less than the effect of number of unique tags, as the transition point was still much higher for StackOverflow given that it should have lowered due to post size.

@@

Another parameter in the random permutation model is the number of rows used to represent all of the different words.
Each word's environment vector is represented by two randomly placed ones and two negative ones across the rows.
All that is required is that the four positions and signs of each of these environment vectors is unique.
But this means multiple environment vectors can overlap when looking at a single row.
This is a lossy form of compression, and allows the number of rows to represent the words to be much less than the total number of words.
As the number of rows decreases, the amount of cross talk between the environment vectors increases, compression increases, and the amount of information stored decreases.
As a third architectural concern, the effect of compression for the random permutation model was examined for Twitter and StackOverflow.

@@

\begin{figure}[!htbp]
  \scalebox{.8}{\resizebox{\linewidth}{!}{\includegraphics{compareMeanDV-acc-dimBySizeSO-crop.pdf}}}
  \caption{
    Effect of compression on random permutation model for StackOverflow.
    Error bars represent the 95\% bootstrapped confidence interval for the mean.
  }
  \label{figContextDimSO}
\end{figure}

@@

\begin{figure}[!htbp]
  \scalebox{.8}{\resizebox{\linewidth}{!}{\includegraphics{compareMeanDV-acc-dimBySizeT-crop.pdf}}}
  \caption{
    Effect of compression on random permutation model for Twitter.
    Error bars represent the 95\% bootstrapped confidence interval for the mean.
  }
  \label{figContextDimT}
\end{figure}

@@

The original BEAGLE model \parencite{Jones2007} actually showed only minimal improvement when incorporating word order as well.
In absolute terms, this model only improved by 2 percentage points in accuracy when trained on the TASA and tested on the TOEFL.
\textcite{Sahlgren2008}'s results for word order for the random permutation model are similar.
At a row size of around \num{2000}, model accuracy improves by less than a percentage point when order information is incorporated.
The results for larger row sizes are somewhat inconsistent, since the full model (order and orderless) sometimes underperforms the order-only model,
which should never happen since the order-only model is a subset of the full model.
\textcite{Sahlgren2008} most likely did not use a regression technique to optimize the weights for each of the order and orderless model terms.
So although incorporating word order into the random permutation model does seem to slightly improve performance,
the strength of the model is more likely to be in how efficiently it can represent the co-occurrence matrix rather than its ability to include word order in the representation.
That is, the strength of the model is in its compressed unordered context component term and not the ordered component, at least for the Twitter dataset studied.

@@

This raises an important applied question: Exactly how much past user behavior is enough for a site like StackOverflow to benefit from adding past user behavior as an additional predictor?
When the model predicts tags from a random sample of StackOverflow users, adding the prior component to the model does not significantly increase model accuracy compared to using context alone.
However, when the model predicts tags from a set of users that ask the most questions on the site, adding the prior does significantly increase accuracy.
For the StackOverflow dataset, a user asks on average 5 questions and adds 3 tags to a post.
So the model uses approximately 15 prior observations when computing the prior component for a randomly-sampled user on the site.
For the popular-users dataset used to generate the results in Figure \ref{figPUAddPrior}, a user asked 180 questions on average.
So for this dataset, the model uses approximately 540 prior observations when computing the prior activation.
So for the StackOverflow site, the prior component begins to generate unique predictive variance above and beyond the context component somewhere between when a user has generated 15 and 540 prior observations.
This result will be different for other sites that have collected a different amount of contextual information.
For sites like StackOverflow where the contextual component is already a strong predictor, it will take a larger number of prior observations before adding the prior component becomes beneficial.
Alternatively for sites like Twitter where contextual information is much more noisy and not as strong of a predictor, it will take fewer observations before adding the prior becomes beneficial.

@@

Given that the storage space required to represent the co-occurrences for the random permutation model is dramatically smaller than the uncompressed Bayesian model,
and that the compression is lossy and essentially random (i.e., no SVD),
it is quite interesting that the random permutation model performs nearly as well as the full Bayesian model for this task.
